Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-04 demystify.tex
updates

\paragraph{Why CS depts do not waive application fee?}  Note that some universities do waive application fee, however most do not.  In any case, application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

@@

By around March you should hear back from most PhD programs you applied to. If you haven't heard back, reach out via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will have to make your decision by around \emph{April 15} (see \href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{why?}).

@@

\paragraph{Low Stipend?} Students often complain their stipend being too low, but it is actually not bad and you can live comfortably with it.  In most cases, it is also enough to support your spouse and kids (many CS PhD students have their family with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at GMU, every year we discuss about improving the benefits, and especially stipend, for our graduate students.

@@

\textbf{How to get RA?} When a professor recruits you, they might offer you an RA right away (e.g., when you apply).  A common scenario is that you first get admitted with TA, and then after a year or two find an adviser to support you with RA. RA is never guaranteed as it depends on the funding situation of the prof. So you should also pay attention to TA, e.g., see if TA is readily available for PhD students in the program.

@@

Fellowship is another type of funding that students can get (e.g., from the university, industries, government).
Fellowships are typically competitive and generous, and gives pretty much all benefits tuition/insurance that a TA/RA has.  Moreover, they often give higher stipend (including summer) and open doors for job opportunities (e.g., internship).

@@

You likely need to apply for external fellowships including those from the US government and tech companies (e.g., Google, Microsoft, Facebook).  Fellowships from US government would likely require US citizenship or residency. % However, tech companies including Google, Microsoft, Facebook have fellowships that international students can apply for.
Prestigious external fellowships typically require a clear and good research plan, so it might be a good idea to wait until at least your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.

@@

\subsection{Selecting and Ranking Schools for CS}\label{sec:selecting-ranking-schools}
\begin{center}
  \includegraphics[scale=0.5]{files/c1.png}
\end{center}

International students not familiar with US universities often put them into \emph{two} bins:  (i) very top schools that they dream about, and (ii) everything else.  In many cases, they use resources such as rankings from US News, which are not very transparent and \emph{highly} questionable\footnote{\url{https://cra.org/cra-statement-us-news-world-report-rankings-computer-science-universities/}}.  Sometimes these students evaluate CS programs using the reputations non-CS programs such as medical, math, or physics.
They even rank universities based on popular states they know in the US, e.g., California and New York.  Clearly, there are so many thing wrong with these methods.

@@

However, in general, rankings can be superficial and you need to do more research to be informed and make better decision. For example, if you get admissions to several places, you should consider attending Open Houses and contact profs. that you're interested in at those place and talk to them. They would be more willing to chat with you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students. See more in \autoref{sec:accepted} on what to do after getting admitted.

@@

\section{Choosing an Adviser}
Obviously, there is no one-size-fit-all answer to this question. The best adviser is the one that you can work well with.  This is difficult and takes time. Fortunately, unlike many countries that require finding an adviser and research topic before starting the PhD (\autoref{sec:non-us-differences}), CS PhD programs in the US will typically give you a couple of years to search for advisers and research topics.  This is especially true if you're admitted with TA (\autoref{sec:ta}), which gives you time to explore and find adviser.

@@

Here are some general advice to find professor.  Assuming you're not familiar with any particular one, then first, search for profs. that share similar research interests.  For example, in CSRankings, if you want to work with PL, you can search for those publish in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.  Then you do research about that professor and contact them (\autoref{sec:contact}).

>>>>

\paragraph{Why CS depts do not waive application fee?}  Note that some universities do waive application fee (e.g., for domestic students (\autoref{sec:domestic-students})), however most do not.  In any case, application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

@@

By around March you should hear back from most PhD programs you applied to. If you haven't heard back, reach out via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will have to make your decision by around \emph{April 15} (\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{why this date?}).

@@

\paragraph{Low Stipend?} Students often complain their stipend being too low. However, in most cases it is actually not bad and you can live comfortably with it.  It might also be enough to support your spouse and kids (many CS PhD students have their family with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at GMU, every year we discuss about improving the benefits, and especially stipend, for our graduate students.

@@

\textbf{How to get RA?} When a professor recruits you, they might offer you an RA right away (so you start with an RA).  However, a more common scenario is that you first get admitted with TA, and then after a year or two find an adviser to support you with RA.

It is important to note that RA is never guaranteed as it depends on the funding situation of the prof. So you should also pay attention to TA, which is a good backup plan (remember, typically TA and RA have similar benefits). This means you should also ask if TA is readily available for PhD students in the program.

@@

Fellowship is another type of funding that students can get (e.g., from the university, industries, government).
Fellowships are typically competitive and generous, and gives pretty much all benefits tuition/insurance that a TA/RA has.  They might even give higher stipend (including summer) and open doors for job opportunities such as internships.

@@

For external fellowships including those from the US government and tech companies (e.g., Google, Microsoft, Facebook), you will need to apply.  Fellowships from US government would likely require US citizenship or residency. % However, tech companies including Google, Microsoft, Facebook have fellowships that international students can apply for.
Prestigious external fellowships typically require a clear and good research plan, so it might be a good idea to wait until at least your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.

@@

\subsection{Selecting and Ranking Schools}\label{sec:selecting-ranking-schools}
\begin{center}
  \includegraphics[scale=0.5]{files/c1.png}
\end{center}

International\footnote{Even domestic students who have done their undergrad in the US might not know much about CS PhD programs in the US.} students often put schools into two bins: (i) very top schools that they dream about, and (ii) everything else.  In many cases, they use rankings from US News, which are not very transparent and highly questionable\footnote{\url{https://cra.org/cra-statement-us-news-world-report-rankings-computer-science-universities/}}.  Sometimes these students evaluate CS programs using the reputations non-CS programs such as medical, math, or physics.
They even rank universities based on popular states they know in the US, e.g., California and New York.  Clearly, there are so many thing wrong with these methods.

@@

However, in general, rankings can be superficial and you need to do more research to be informed and make better decision. For example, if you get admissions to several places, you should consider attending Open Houses (\autoref{sec:accepted}) and contact profs. that you're interested in at those place and talk to them. They would be more willing to chat with you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students. See more in \autoref{sec:accepted} on what to do after getting admitted.

@@

\section{Choosing an Adviser}
Obviously, there is no one-size-fit-all answer to this question. The best adviser is the one that you can work well with.  But how do you find such a person?

Fortunately, unlike many countries that require finding an adviser and research topic before starting the PhD (\autoref{sec:non-us-differences}), CS PhD programs in the US will typically give you a couple of years to search for advisers and research topics.  This is especially true if you're admitted with TA (\autoref{sec:ta}), which gives you time to explore and find adviser.

@@

Here are some general advice to find professor.  Assuming you're not familiar with any particular one, then first search for profs. that share similar research interests.  For example, in CSRankings, if you want to work with PL, you can search for those publish in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.  Then you do research about that professor and contact them (\autoref{sec:contact}).
